
## Read data

```{r cars}
data = read.csv("days.csv")
```

## Fit multiple linear regression model

```{r pressure, echo=FALSE}
data$season<-as.factor(data$season)
data$yr<-as.factor(data$yr)
data$mnth<-as.factor(data$mnth)
data$holiday<-as.factor(data$holiday)
data$weekday<-as.factor(data$weekday)
data$weathersit<-as.factor(data$weathersit)





model = lm(cnt ~ season + yr + mnth + holiday + weekday+ workingday+weathersit+temp+atemp+hum+windspeed, data = data)
summary(model)
```
.8484 of the variation in count can be explained by the predictors.

The most significant variables are windspeed, weathersit, and humidity. 

## Let's check for correlation between independent variables in the model

```{r}
# install.packages('GGally')
library(GGally)
data_x = data[3:13]
ggpairs(data_x, aes( alpha = 0.2), lower=list(combo=wrap("facethist",  binwidth=0.5)))
```

We notice that atemp and temp and highly correlated

## Let's check for multicolinearity

```{r error=TRUE}
library(car)
vif(model)
```

The VIF function throws an error - this suggests some perfect multicolinearity. 
Let's see what variable is causing the issue.

```{r}
ld.vars <- attributes(alias(model)$Complete)$dimnames[[1]]
ld.vars
```

Workingday is linearly dependent variable. Let's try VIF after removing the variable.

```{r}
model2 = lm(cnt ~ season + yr + mnth + holiday + weekday+weathersit+temp+atemp+hum+windspeed, data = data)
```
```{r}
vif(model2)
```
VIF values larger than 10 are a sign for high, not tolerable correlation of model predictors, therefore season, mnth, temp, and atemp are poor predictors to use.
