---
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancypagestyle{plain}{\pagestyle{fancy}}
- \headheight 35pt
- \fancyhead[LE,LO]{Abowath, Blakely, \\ Garrison, Loui, Peir \\}
- "\\fancyhead[CO,CE]{\\textbf{\\Large Forecasting Bikesharing Usage \\\\ Team 10
  - Final Report} \\\\ https://github.gatech.edu/MGT-6203-Fall-2023-Canvas/Team-10}"
output:
  html_document:
    df_print: paged
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
require(gridExtra)
library(kableExtra)
library(broom)
```
# Forecasting Bikesharing Usage for DC's Capital Bikshare System

# Table of Contents

* Introduction
* Overview of Project
* Literature Review Summary
* Overview of Data
* Overview of Modeling
* Conclusion
* Works Cited

# Introduction

Bikesharing systems are an increasingly popular solution in major urban areas to increase the usage of bicycles as a mode of transport. The riding of bikes helps to improve the lives of both users, as well as non-users, as each bike trip potentially represents a trip that would otherwise have required a car. We used data from the DC Capital Bikeshare from 2011 to 2017. 

The purpose of this analysis is to determine variables/factors that help estimate bike usage and develop a model that forecasts the usage based on certain predictor variables.

# Overview of Project

The idea of the project was to use different data acquisition, cleaning and modeling techniques to determine if we could forecast the bike usage. If we are able to model and forecast the bike usage, then it would allow the company to plan for the best route to increasing our fleet as well as expanding the available stations that are offered.

Some of the questions were: can we use the data to know when our usage is lower to repair bikes? When should we start increasing our fleet to best meet demand? Does weather and seasons have an impact on our business model? 

We made an initial hypothesis that we would see a higher usage during the summer months and when the weather was nicer. With our initial exploratory data analysis, we could see that there is a trend and seasonality to the usage. The question then became, can we model this through a time series model. Also could we see what features were key factors in determining bike usage. Does the season or weather have an affect?

# Literature Review Summary

We reviewed a few papers modeling bikeshare using in different cities across the globe
(see Works Cited). These papers generally shared the same findings, which largely align
with our Capital Bikeshare data:

* Usage increases as temperature increases, then starts to decrease as temperatures go into the 90s (Fahrenheit)
* Precipitation of any amount discourages cycling
* High winds can have a negative effect on cycling
* Usage is often higher in spring and summer, and lowest in winter

# Overview of Data

## Initial Data Set:

We started by just looking at two years of Capital Bikeshare usage from this dataset:
https://archive.ics.uci.edu/dataset/275/bike+sharing+dataset 

The dataset contains 2011 and 2012 historical usage data from Washington, DC’s public 
Capital Bikeshare program, one of the first large scale 
bikeshare programs in the nation. Usage data is broken out by day and by hour.
Additional data included a variety of information on weather, season, and whether
a day was a holiday.

## Data Cleaning Process

The dataset had required minimal cleaning. We had to convert several variables into factor variables (`season`, `holiday`, `weekday`, `workingday`, `weather`). Additionally we noted that the key for our dataset mislabeled the season variable, which was trivial to correct. Fortunately, there was no missing data.

## Additional Scraping, Cleaning

Initial EDA on the dataset we began with indicated we did not have sufficient
data to fit any models that could improve upon linear regression. So, we scraped
and combined two different datasets. We used python scripts located in the
Other Resources directory to do this. 

Specifically, we ran get_bikeshare_data.py to get bikeshare data from Capital Bikeshare directly, 
then we ran join_data.py (which imports from noaa.py) to join the data 
with weather data from the NOAA API. 

We used the same process to add the additional weather columns 
to the 2011 and 2012 as well.

### Sources for additional datasets:

* Capital Bikeshare usage data from 01-01-2013 through 12-31-2017, from
Capital Bikeshare (https://ride.capitalbikeshare.com/system-data)
* Weather data for DC for the same time period, from NOAA (https://www.ncdc.noaa.gov/cdo-web/webservices/v2)

The Bikeshare data only counted trips taken, and did not distinguish between
registered or casual users.

The Weather data included:
* TMAX: high temperature (in tenths of degree Celsius), 
* PRCP: precipitation (tenths of mm), and 
* AWND: average daily windspeed (km/h). 

The final merged dataset used for analysis is final_df.csv located in the Data folder.

Our R code applies a season variable to the dataset, based on the date.

## Exploratory Data Analysis

```{r usage_hist, echo=F, out.width="60%", fig.align="center", message=FALSE}
library(tidyverse)
library(ggplot2)
final_df <- read_csv('../Data/final_df.csv')
final_df %>% ggplot() + geom_histogram(mapping = aes(cnt, fill="Total"), alpha = 1) +
    labs(
    x = "Bike Usage per Day",
    y = "Density",
    title = "Bikeshare Usage Per Day"
  )
```
Fortunately, Bikeshare usage per day is not heavily skewed

```{r temp_hist, echo=F, out.width="60%", fig.align="center", message=FALSE}

final_df %>% ggplot() + geom_histogram(mapping = aes(TMAX, fill="TMAX"), alpha = 1) +
    labs(
    x = "Daily High Temperature in Tenths of Degree Celsius",
    y = "Density",
    title = "Temperature Distribution"
  )
```

```{r precp_hist, echo=F, out.width="60%", fig.align="center", message=FALSE}

final_df %>% ggplot() + geom_histogram(mapping = aes(PRCP, fill="PRCP"), bins = 100, alpha = 1) +
    labs(
    x = "Precipitation in mm",
    y = "Density",
    title = "Precipitation Distribution"
  )
```

```{r windspeed, echo=F, out.width="60%", fig.align="center", message=FALSE}

final_df %>% ggplot() + geom_histogram(mapping = aes(AWND, fill="AWND"), bins = 100, alpha = 1) +
    labs(
    x = "Windspeed in km/h",
    y = "Density",
    title = "Windspeed Distribution"
  )
```

We see that windspeed and precipitation exhibit rightward skew, which makes sense:
* Most days have no precipitation
* Most days have moderate windspeed, while a few have very high windspeed

By contrast, temperature has leftward skew, with a smaller number of colder days

```{r weekday_usage, echo=F, out.width="75%", fig.align="center"}
final_df %>% mutate(season = as_factor(season)) %>%  
  mutate(day_of_week = as_factor(day_of_week)) %>%
  ggplot(mapping = aes(day_of_week, cnt, color = season)) +
  geom_boxplot() +
  labs(
    x = "Weekday",
    y = "Total Bike Usage",
    title = "Total Bike Usage",
    subtitle = "Given Weekday and Season"
  )
```
We can see in the above chart the effect of season (1 = Winter, 2 = Spring, 3 = Summer,  4= Fall)
as well as the effects of the day of the week on total bikeshare usage. Usage is lowest in the winter,
highest in the summer, with spring and fall in between.

Fortunately, there is not a huge amount of variation in usage between weekdays and weekends.

```{r weekday_usage2, echo=F, out.width="75%", fig.align="center"}
final_df %>% mutate(season = as_factor(season)) %>%  
  mutate(day_of_week = as_factor(day_of_week)) %>%
  ggplot(mapping = aes(dteday, cnt, color = season)) +
  geom_point() +
  labs(
    x = "Date",
    y = "Total Bike Usage",
    title = "Total Bike Usage",
    subtitle = "Total Bike Usage Across All Times in Combined Dataset"
  )
```
The most challenging aspect of our dataset is that usage was not static - it grew
from 2011 through 2017. While this was obviously good for Capital Bikeshare, it meant
that models would need to take into account not just variation within a year,
but overall increasing usage over time.

```{r temp_usage, echo=F, out.width="75%", fig.align="center"}
final_df %>% mutate(season = as_factor(season)) %>%  
  mutate(day_of_week = as_factor(day_of_week))
  ggplot(mapping = aes(TMAX, cnt, color = season)) +
  geom_point() +
  labs(
    x = "TMAX",
    y = "Total Bike Usage",
    title = "Total Bike Usage",
    subtitle = "Total Bike Usage vs. Temperature and Season"
  )
```
Consistent with our literature review, see that as temperature (TMAX) increases, usage
tends to increase, until around 30 degrees celsius, after which usage tends to decrease.

```{r prcp_usage, echo=F, out.width="75%", fig.align="center"}
final_df %>% mutate(season = as_factor(season)) %>%  
  mutate(day_of_week = as_factor(day_of_week)) %>%
  ggplot(mapping = aes(PRCP, cnt, color = season)) +
  geom_point() +
  labs(
    x = "PRCP",
    y = "Total Bike Usage",
    title = "Total Bike Usage",
    subtitle = "Total Bike Usage vs. Precipitation and Season"
  )
```
The relationship between precepitation and usage appears somewhat weaker than temperature.
Still, the highest precipitation days tend not to have high usage, consitent with an overall negative effect of preciptiation on usage.

```{r awnd_usage, echo=F, out.width="75%", fig.align="center"}
final_df %>% mutate(season = as_factor(season)) %>%  
  mutate(day_of_week = as_factor(day_of_week)) %>%
  ggplot(mapping = aes(AWND, cnt, color = season)) +
  geom_point() +
  labs(
    x = "AWND",
    y = "Total Bike Usage",
    title = "Total Bike Usage",
    subtitle = "Total Bike Usage vs. Windspeed and Season"
  )
```
We can also see a slightly negative relationship between windspeed and usage.

A correlation matrix comparing our numeric variables validates our graphical EDA:
* There is a positive correlation between temperature and count
* There are negative correlations between precipitation and count, and windspeed and count

We can also see the negative correlation of temperature with windspeed. Fortunately,
precipitation has minimal correlation with temperature (-0.006) and relatively low
correlation with windspeed (0.093).

```{r correlation_matrix, echo=F, out.width="75%", fig.align="center"}
final_df %>% select(TMAX, PRCP, AWND, cnt) %>% cor()
```

## Key Predictors

# Overview of Modeling

For our modeling, we used the `final_df.csv` file in the Other Resources directory, which is described in the accompanying readme.

## Model Types and Comparison

### Linear Regression
```{r linreg, echo=F, out.width="75%", fig.align="center"}
final_df <- final_df %>% mutate(season = as_factor(season)) %>%  
  mutate(day_of_week = as_factor(day_of_week))
lin_reg <- lm(cnt ~ TMAX + PRCP + AWND + day_of_week + season, data = final_df)
summary(lin_reg)
pred <- predict(lin_reg, final_df)
plot(final_df$cnt)
lines(pred, col = 'red')
```
A naive linear regression does not incorporate any time series factors, has an adjusted R-squared of 0.45. We can see that the linear regression over-predicts usage in early years,
and under-predicts usage in later years, due to lacking a trend/time series component.

## Model Selection and Optimization

## Model Performance

## Detailed evaluation and interpretation of results

#Conclusion

One finding was that forecasting effectively requires more data than two years of usage. Our initial models trained just on 2011-2012 data did not greatly outperform linear regression.

# Works Cited

Bean, R., Pojani, D., & Corcoran, J. (2021). How does weather affect bikeshare use? A comparative analysis of forty cities across climate zones. _Journal of Transport Geography_, 95. https://doi.org/10.1016/j.jtrangeo.2021.103155.

Eren, E., & Uz, V. E. (2020). A review on bike-sharing: The factors affecting bike-sharing demand. _Sustainable Cities and Society_, 54. https://doi.org/10.1016/j.scs.2019.101882

Ashgar, H. I., Elhenawy, M., & Rakha, H. A. (2019). Modeling bike counts in a bike-sharing system considering the effect of weather conditions. _Case Studies on Transport Policy_, 7(2), 261-268. https://doi.org/10.1016/j.cstp.2019.02.011
